The Clinicopathological Features and Survival Outcomes of Patients with Different Metastatic Sites in Stage IV Endometrial Cancer: A SEER-Based Cohort Study

### 2025-03-14 code/eda-code/edamz.qmd
part 3 project v1

```{r}
# Summary of Age recode variable (categorical)
p1 <- mydata %>%
  group_by(`Age recode with <1 year olds`) %>%
  summarise(Count = n())

@@

```{r}
# Bar plot for Age recode distribution (categorical)
p2 <- mydata %>%
  ggplot(aes(x = `Age recode with <1 year olds`)) + 
  geom_bar(fill = "skyblue") + 
  labs(title = "Age Distribution (Categorical)", x = "Age Group", y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) # Rotate x-axis labels for readability

@@

```{r}
# Summary of Race recode variable (categorical)
# - Groups the dataset by the "Race recode (W, B, AI, API)" variable.
# - Counts the number of observations in each race category.
p3 <- mydata %>%
  group_by(`Race recode (W, B, AI, API)`) %>%
  summarise(Count = n())

@@

# Save the summary table
# - Stored in "results/edatables/" with the prefix "eda" to indicate exploratory analysis.
summarytable_file = here("results", "edatables", "edaracesummary.rds")
saveRDS(p3, file = summarytable_file)
```
```{r}
# Bar plot for Race recode distribution (categorical)
# - Visualizes the distribution of race categories.
# - Uses "skyblue" fill for better readability.
p4 <- mydata %>%
  ggplot(aes(x = `Race recode (W, B, AI, API)`)) + 
  geom_bar(fill = "skyblue") + 
  labs(title = "Race Distribution", x = "Race Category", y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) # Rotates x-axis labels for better readability

@@

```
Now exploring Year of Diagnosis 
```{r}
# Summary of Year of Diagnosis variable (categorical)
# - Groups the dataset by "Year of diagnosis" and counts occurrences.
p5 <- mydata %>%
  group_by(`Year of diagnosis`) %>%
  summarise(Count = n())

@@

```
```{r}
# Bar plot for Year of Diagnosis distribution (categorical)
# - Visualizes the distribution of diagnosis years.
# - Uses "skyblue" fill for better readability.
p6 <- mydata %>%
  ggplot(aes(x = factor(`Year of diagnosis`))) +  # Treat as categorical
  geom_bar(fill = "skyblue") + 
  labs(title = "Year of Diagnosis Distribution", x = "Year", y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) # Rotates x-axis labels for better readability

@@

```{r}
# Summary of SEER Combined Mets at DX variables (categorical)
# - Groups data by metastasis categories and counts occurrences.
p7 <- mydata %>%
  summarise(
    Bone_Mets = sum(`SEER Combined Mets at DX-bone (2010+)` == "Yes", na.rm = TRUE),
    Brain_Mets = sum(`SEER Combined Mets at DX-brain (2010+)` == "Yes", na.rm = TRUE),
    Liver_Mets = sum(`SEER Combined Mets at DX-liver (2010+)` == "Yes", na.rm = TRUE),
    Lung_Mets = sum(`SEER Combined Mets at DX-lung (2010+)` == "Yes", na.rm = TRUE),
      ) %>%
  pivot_longer(cols = everything(), names_to = "Metastasis_Site", values_to = "Count")

@@

```{r}
# Summary of SEER Cause-Specific Death Classification variable (categorical)
# - Counts the number of occurrences for each cause-specific death classification category.
# - Also calculates the percentage for each category.
p9 <- mydata %>%
  group_by(`SEER cause-specific death classification`) %>%
  summarise(Count = n(), .groups = "drop") %>%
  mutate(Percentage = round((Count / sum(Count)) * 100, 1))

@@

```{r}
# Bar plot for SEER Cause-Specific Death Classification distribution
# - Visualizes the distribution of cause-specific death classification.
p10 <- p9 %>%
  ggplot(aes(x = `SEER cause-specific death classification`, y = Count, fill = `SEER cause-specific death classification`)) + 
  geom_bar(stat = "identity") + 
  labs(title = "SEER Cause-Specific Death Classification", x = "Cause-Specific Death Category", y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  guides(fill = "none")  # Removes legend since x-axis labels already describe the categories

@@

```{r}
# Summary of Survival Months variable (continuous)
# - Computes descriptive statistics for survival months.
p11 <- mydata %>%
  summarise(
    Min_Survival = min(`Survival months`, na.rm = TRUE),
    Max_Survival = max(`Survival months`, na.rm = TRUE),
    Mean_Survival = mean(`Survival months`, na.rm = TRUE),
    Median_Survival = median(`Survival months`, na.rm = TRUE),
    SD_Survival = sd(`Survival months`, na.rm = TRUE),
    Count = n()
  )

@@

```
```{r}
# Histogram for Survival Months distribution
# - Visualizes the distribution of survival months.
p12 <- mydata %>%
  ggplot(aes(x = `Survival months`)) + 
  geom_histogram(binwidth = 5, fill = "skyblue", color = "black") + 
  labs(title = "Survival Months Distribution", x = "Survival Months", y = "Count") +
  theme_minimal()

>>>>

```{r}
# Summary of Age recode variable (categorical)
p1 <- mydata %>%
  group_by(`age_group`) %>%
  summarise(Count = n())

@@

```{r}
# Bar plot for Age recode distribution (categorical)
p2 <- mydata %>%
  ggplot(aes(x = `age_group`)) + 
  geom_bar(fill = "skyblue") + 
  labs(title = "Age Distribution (Categorical)", x = "Age Group", y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) # Rotate x-axis labels for readability

@@

```{r}
# Summary of Race recode variable (categorical)
# - Groups the dataset by the "Race recode (W, B, AI, API)" variable.
# - Counts the number of observations in each race category.
p3 <- mydata %>%
  group_by(`race`) %>%
  summarise(Count = n())

@@

# Save the summary table
# - Stored in "results/edatables/" with the prefix "eda" to indicate exploratory analysis.
summarytable_file = here("results", "edatables", "edaracesummary.rds")
saveRDS(p3, file = summarytable_file)
```
```{r}
# Bar plot for Race recode distribution (categorical)
# - Visualizes the distribution of race categories.
# - Uses "skyblue" fill for better readability.
p4 <- mydata %>%
  ggplot(aes(x = race)) + 
  geom_bar(fill = "skyblue") + 
  labs(title = "Race Distribution", x = "Race Category", y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) # Rotates x-axis labels for better readability

@@

```
Now exploring Year of Diagnosis 
```{r}
# Summary of Year of Diagnosis variable (categorical)
# - Groups the dataset by "Year of diagnosis" and counts occurrences.
p5 <- mydata %>%
  group_by(diagnosis_year) %>%
  summarise(Count = n())

@@

```
```{r}
# Bar plot for Year of Diagnosis distribution (categorical)
# - Visualizes the distribution of diagnosis years.
# - Uses "skyblue" fill for better readability.
p6 <- mydata %>%
  ggplot(aes(x = factor(diagnosis_year))) +  # Treat as categorical
  geom_bar(fill = "skyblue") + 
  labs(title = "Year of Diagnosis Distribution", x = "Year", y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) # Rotates x-axis labels for better readability

@@

```{r}
# Summary of SEER Combined Mets at DX variables (categorical)
# - Groups data by metastasis categories and counts occurrences.
p7 <- mydata %>%
  summarise(
    Bone_Mets = sum(mets_bone == "Yes", na.rm = TRUE),
    Brain_Mets = sum(mets_brain == "Yes", na.rm = TRUE),
    Liver_Mets = sum(mets_liver == "Yes", na.rm = TRUE),
    Lung_Mets = sum(mets_lung == "Yes", na.rm = TRUE),
      ) %>%
  pivot_longer(cols = everything(), names_to = "Metastasis_Site", values_to = "Count")

@@

```{r}
# Summary of SEER Cause-Specific Death Classification variable (categorical)
# - Counts the number of occurrences for each cause-specific death classification category.
# - Also calculates the percentage for each category.
p9 <- mydata %>%
  group_by(cause_specific_death) %>%
  summarise(Count = n(), .groups = "drop") %>%
  mutate(Percentage = round((Count / sum(Count)) * 100, 1))

@@

```{r}
# Bar plot for SEER Cause-Specific Death Classification distribution
# - Visualizes the distribution of cause-specific death classification.
p10 <- p9 %>%
  ggplot(aes(x =cause_specific_death, y = Count, fill = cause_specific_death)) + 
  geom_bar(stat = "identity") + 
  labs(title = "SEER Cause-Specific Death Classification", x = "Cause-Specific Death Category", y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  guides(fill = "none")  # Removes legend since x-axis labels already describe the categories

@@

```{r}
# Summary of Survival Months variable (continuous)
# - Computes descriptive statistics for survival months.
p11 <- mydata %>%
  summarise(
    Min_Survival = min(survival_months, na.rm = TRUE),
    Max_Survival = max(survival_months, na.rm = TRUE),
    Mean_Survival = mean(survival_months, na.rm = TRUE),
    Median_Survival = median(survival_months, na.rm = TRUE),
    SD_Survival = sd(survival_months, na.rm = TRUE),
    Count = n()
  )

@@

```
```{r}
# Histogram for Survival Months distribution
# - Visualizes the distribution of survival months.
p12 <- mydata %>%
  ggplot(aes(x = survival_months)) + 
  geom_histogram(binwidth = 5, fill = "skyblue", color = "black") + 
  labs(title = "Survival Months Distribution", x = "Survival Months", y = "Count") +
  theme_minimal()

### 2025-03-14 code/processing-code/processingcodemz.qmd
part 3 project v1

@@

```{r}
processeddata <- d6  # Assign cleaned dataset
```

@@

# Notes

Removing individuals with "faulty" or missing data is one approach. However, depending on the question and analysis approach, different cleaning methods might be better (e.g., keeping individuals with some missing information).

>>>>

## Excluding Patients with Unknown metastasis

```{r}
d7 <- d6 %>% dplyr::filter(
    # Exclude patients where all four metastasis indicators are "Unknown"
    !(`SEER Combined Mets at DX-bone (2010+)` == "Unknown" &
      `SEER Combined Mets at DX-brain (2010+)` == "Unknown" &
      `SEER Combined Mets at DX-liver (2010+)` == "Unknown" &
      `SEER Combined Mets at DX-lung (2010+)` == "Unknown")
)

# Check the updated dataset after filtering
skimr::skim(d7)

```

## Renaming all the variables for the ease of analysis 
```{r}
d8 <- d7 %>% dplyr::rename(
    # Demographic Information
    age_group = `Age recode with <1 year olds`,  # Age category of the patient
    sex = Sex,  # Patient's sex
    race = `Race recode (W, B, AI, API)`,  # Race classification
    ethnicity = `Origin recode NHIA (Hispanic, Non-Hisp)`,  # Ethnicity classification
    race_ethnicity = `Race and origin recode (NHW, NHB, NHAIAN, NHAPI, Hispanic)`,  # Combined race & ethnicity

    # Diagnosis & Cancer Staging
    diagnosis_year = `Year of diagnosis`,  # Year of cancer diagnosis
    cancer_stage = `Derived AJCC Stage Group, 7th ed (2010-2015)`,  # Cancer staging classification
    surgery_primary_site = `RX Summ--Surg Prim Site (1998+)`,  # Surgery performed at the primary site

    # Metastasis Indicators
    mets_bone = `SEER Combined Mets at DX-bone (2010+)`,  # Metastasis at bone
    mets_brain = `SEER Combined Mets at DX-brain (2010+)`,  # Metastasis at brain
    mets_liver = `SEER Combined Mets at DX-liver (2010+)`,  # Metastasis at liver
    mets_lung = `SEER Combined Mets at DX-lung (2010+)`,  # Metastasis at lung
    mets_distant_ln = `Mets at DX-Distant LN (2016+)`,  # Metastasis at distant lymph nodes

    # Cause of Death & Survival
    cause_of_death = `COD to site recode`,  # Cause of death classification
    cause_specific_death = `SEER cause-specific death classification`,  # SEER-specific cause of death
    other_cause_of_death = `SEER other cause of death classification`,  # Other causes of death
    survival_months = `Survival months`,  # Number of survival months
    survival_flag = `Survival months flag`,  # Flag indicating survival data availability
    vital_status = `Vital status recode (study cutoff used)`,  # Vital status (Alive/Dead)
    cod_km = `COD to site rec KM`,  # Cause of death classification for Kaplan-Meier

    # Socioeconomic & Patient ID
    patient_id = `Patient ID`,  # Unique patient identifier
    median_income = `Median household income inflation adj to 2019`,  # Adjusted median income
    rural_urban_code = `Rural-Urban Continuum Code`  # Rural-Urban classification
)

# Check new variable names to ensure proper renaming
colnames(d8)

```

@@

```{r}
processeddata <- d8  # Assign cleaned dataset
```

@@

### 2025-03-14 products/manuscript/Manuscript.qmd
part 3 project v1

---
title: "The Clinicopathological Features and Survival Outcomes of Patients with Different Metastatic Sites in Stage IV Endometrial Cancer:A SEER-Based Cohort Study"
author: "Mohammed Zuber"
format:
  docx:
    toc: false
    number-sections: true
    highlight-style: github
bibliography: ../../assets/dataanalysis-references.bib
csl: ../../assets/american-journal-of-epidemiology.csl
editor: 
  markdown: 
    wrap: sentence
---

The structure below is one possible setup for a manuscript, or a general data analysis project (including the course project).
Adjust as needed.
You don't need to have exactly these sections, but the content covering those sections should be addressed.

This uses MS Word as output format.
[See here](https://quarto.org/docs/output-formats/ms-word.html) for more information.
You can switch to other formats, like html or pdf.
See [the Quarto documentation](https://quarto.org/) for other formats.

@@

The analysis aims to identify survival differences based on metastatic patterns, demographic disparities, socioeconomic influences, and temporal trends in prognosis.
Kaplan-Meier survival analysis will used to compare the survival among the study cohorts.
The incidence of metastasis will be calculated in stage IV endometrial cancer.
A Cox regression model will be performed to compute the adjusted Hazard Ratios (aHR) and 95% confidence intervals (CI).

@@

## Basic statistical analysis

*To get some further insight into your data, if reasonable you could compute simple statistics (e.g. simple models with 1 predictor) to look for associations between your outcome(s) and each individual predictor variable. Though note that unless you pre-specified the outcome and main exposure, any "p\<0.05 means statistical significance" interpretation is not valid.*

@fig-result shows a scatterplot figure produced by one of the R scripts.

```{r}
#| label: fig-result
#| fig-cap: "Height and weight stratified by gender."
#| echo: FALSE
knitr::include_graphics(here("results","figures","height-weight-stratified.png"))
```

>>>>

---
title: "The Clinicopathological Features and Survival Outcomes of Patients with Different Metastatic Sites in Stage IV Endometrial Cancer: A SEER-Based Cohort Study"
author: "Mohammed Zuber"
format:
  docx:
    toc: false
    number-sections: true
    highlight-style: github
bibliography: ../../assets/dataanalysis-references.bib
csl: ../../assets/american-journal-of-epidemiology.csl
editor: 
  markdown: 
    wrap: sentence
---

@@

The analysis aims to identify survival differences based on metastatic patterns, demographic disparities, socioeconomic influences, and temporal trends in prognosis.
Kaplan-Meier survival analysis will used to compare the survival among the study cohorts.
The incidence of metastasis will be calculated in stage IV endometrial cancer.
In the bivariate analysis, we fitted separate Cox regression models for each variable, including metastasis status (bone, brain, liver, and lung), race, age group, year of diagnosis, and median household income, to estimate their individual hazard ratios (HRs).
For the multivariate analysis, we constructed a Cox model incorporating all covariates simultaneously to adjust for potential confounders and determine independent predictors of survival.
A Cox regression model was performed to compute the adjusted Hazard Ratios (aHR).

@@

## Basic statistical analysis
### Bivariate Analysis: Metastasis and Survival

In the bivariate Cox proportional hazards analysis, all four metastasis types were significantly associated with an increased risk of mortality.
Bone metastasis was associated with a 56% higher hazard of death (HR = 1.56), while brain metastasis showed the highest risk, nearly a 92% increased hazard of death (HR = 1.92).
Liver metastasis was also strongly associated with worse survival, showing an 81% higher hazard of death (HR = 1.81), and lung metastasis increased the hazard of death by 63% (HR = 1.63).
These findings indicate that the presence of metastases, particularly in the brain and liver, is a strong predictor of poor survival outcomes.
@tbl-summaryhrmetastasis shows the HR for different metastatis sites

```{r}
#| label: tbl-summaryhrmetastasis
#| tbl-cap: "Proportional hazards analysis for different metastasis."
#| echo: FALSE
resulttable=readRDS(here("results", "tables", "mets_HR.rds"))
knitr::kable(resulttable)
```

@fig-km-curve-bone-mets illustrates the survival probability of patients with and without bone metastasis over time.
The red curve represents patients without bone metastasis, while the blue curve represents those with bone metastasis.
Patients with bone metastasis have significantly worse survival outcomes, as indicated by the lower survival probability throughout the follow-up period.

```{r}
#| label: fig-km-curve-bone-mets
#| fig-cap: "Kaplan-Meier survival curves for Bone Metastasis"
#| echo: FALSE
knitr::include_graphics(here("results", "figures", "KM_curve_bone_mets.png"))
```

@fig-km-curve-brain-mets shows that patients with brain metastasis have significantly lower survival probabilities than those without.
The log-rank test (p \< 0.0001) confirms a statistically significant difference, indicating that brain metastasis is associated with worse survival outcomes.

```{r}
#| label: fig-km-curve-brain-mets
#| fig-cap: "Kaplan-Meier survival curves for Brain Metastasis"
#| echo: FALSE
knitr::include_graphics(here("results", "figures", "KM_curve_brain_mets.png"))
```

@fig-km-curve-liver-mets indicates that patients with liver metastasis (blue) have significantly lower survival probabilities than those without (red).
The log-rank test (p \< 0.0001) confirms a statistically significant difference, suggesting that liver metastasis is associated with poorer survival outcomes.

```{r}
#| label: fig-km-curve-liver-mets
#| fig-cap: "Kaplan-Meier survival curves for Liver Metastasis"
#| echo: FALSE
knitr::include_graphics(here("results", "figures", "KM_curve_liver_mets.png"))
```

@fig-km-curve-lung-mets shows that patients with lung metastasis (blue) have significantly lower survival probabilities than those without (red).
The log-rank test (p \< 0.0001) confirms a statistically significant difference, indicating that lung metastasis is associated with worse survival outcomes.

```{r}
#| label: fig-km-curve-lung-mets
#| fig-cap: "Kaplan-Meier survival curves for Lung Metastasis"
#| echo: FALSE
knitr::include_graphics(here("results", "figures", "KM_curve_lung_mets.png"))
```
### Bivariate Analysis: Race and Survival
Patients identified as  had significantly worse survival outcomes than White patients, with a 37% higher risk of death (HR = 1.37, p < 0.0001). In contrast, Asian patients had a lower risk of death compared to White patients (HR = 0.86, p = 0.0076), suggesting a potential survival advantage. The survival difference between American Indian and White patients was not statistically significant (HR = 1.21, p = 0.29), indicating no clear evidence of an increased or decreased risk. @tbl-summaryhrrace shows the HR for patients with different race.
```{r}
#| label: tbl-summaryhrrace
#| tbl-cap: "Proportional hazards analysis for different Race."
#| echo: FALSE
resulttable=readRDS(here("results", "tables", "race_HR.rds"))
knitr::kable(resulttable)
```
### Bivariate Analysis: Age and Survival
The hazard ratio (HR) estimates for different age groups indicate a general trend of increasing risk with age, though statistical significance varies. Compared to the reference group (20-24 years), younger age groups (25-74 years) show HRs between 3.5 and 6.2, but their p-values are above 0.05, suggesting no statistically significant difference in survival risk. However, individuals aged 80-84 years (HR = 7.99, p = 0.038) and 85+ years (HR = 11.01, p = 0.017) have significantly higher mortality risk, indicating poorer survival outcomes in older patients. These findings suggest that age is an important predictor of survival, with the highest risk observed in the oldest age group. @tbl-summaryhrage shows the HR for diferrent age groups.
```{r}
#| label: tbl-summaryhrage
#| tbl-cap: "Proportional hazards analysis for different age group."
#| echo: FALSE
resulttable=readRDS(here("results", "tables", "age_HR.rds"))
knitr::kable(resulttable)
```
### Multivariable analysis

The multivariable Cox proportional hazards model results indicate that metastasis to the bone (HR = 1.30, p < 0.001), brain (HR = 1.72, p < 0.001), liver (HR = 1.66, p < 0.001), and lung (HR = 1.51, p < 0.001) are all significantly associated with poorer survival outcomes. Among racial groups, Black patients (HR = 1.39, p < 0.001) had significantly worse survival compared to White patients, while Asian or Pacific Islander and American Indian/Alaska Native groups did not show a statistically significant difference.

Regarding age, survival risk generally increased with older age, with statistically significant increases observed in the 70-74 years (HR = 7.26, p = 0.048), 75-79 years (HR = 7.91, p = 0.039), 80-84 years (HR = 9.53, p = 0.025), and 85+ years (HR = 12.75, p = 0.011) groups, indicating poorer survival outcomes in older patients.

Year of diagnosis did not show a consistent or significant trend, suggesting no clear improvement or decline in survival based on the year of diagnosis. Household income was associated with survival, with patients in the $50,000 - $54,999 (HR = 1.16, p = 0.011) and $65,000 - $69,999 (HR = 1.12, p = 0.030) categories showing significantly higher risk compared to those earning $75,000+. However, other income groups did not show significant differences.

Overall, metastasis, age, and race appear to be key determinants of survival, with Black race and older age groups associated with poorer outcomes.
